---
title: Clean and preprocess the raw data to a dataset suitable for linear mixed modelling.
author: "Kenneth Wong Chi Yin"
date: "06/05/2024
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


### Read the raw dataset
```{r, results='hide'}
VERSION="paper20231025"
BATCHID="CPH.20211217"
DATAPATH=paste0("/home/kenneth/data/metabolic/ver20220920/")
OUTPATH=paste0("/home/kenneth/out/metabolic/", BATCHID, "/", VERSION, "/phenotype/")

# Read input dataset
df.dose <- read.csv(paste0(DATAPATH, "medication_dose_calculated.csv"))
df.clinic <- read.csv(paste0(DATAPATH, "metabolic_cleaned.csv"))

# Remove invalid records
rm_idx <- c()
df.clinic <- df.clinic[df.clinic$ID!=487,]  # Gender= "\"
idx <- which(df.clinic$ID==584&df.clinic$BL_Date=="19/8/2019"&is.na(df.clinic$LDL_Chol))
rm_idx <- c(rm_idx, idx)
idx <- which(df.clinic$ID==274&df.clinic$Date=="1/12/2010"&df.clinic$baseline==0)
rm_idx <- c(rm_idx, idx)
idx <- which(df.clinic$ID==419&df.clinic$Date=="22/12/2011")[1]
rm_idx <- c(rm_idx, idx)
idx <- which(df.clinic$ID==545&df.clinic$Date=="1/8/2011")[1]
rm_idx <- c(rm_idx, idx)
idx <- which(df.clinic$ID==701&df.clinic$Date=="29/2/2016")[1]
rm_idx <- c(rm_idx, idx)
df.clinic <- df.clinic[-rm_idx,]

# Convert education from character to integer
df.clinic$education <- as.integer(df.clinic$education)

# Classification of drugs
# Candidate SGA drugs for variable selection
SGA_drugs <- c("CLOZAPINE","OLANZAPINE","ARIPIPRAZOLE","AMISULPRIDE","PALIPERIDONE","RISPERIDONE","QUETIAPINE", "ZIPRASIDONE", "LURASIDONE")

# Candidate FGA drugs for variable selection
FGA_drugs <- c("CHLORPROMAZINE","FLUPHENAZINE", "HALOPERIDOL", "PERPHENAZINE", "TRIFLUOPERAZINE", "PROCHLORPERAZINE", "PIMOZIDE")

# Candidate metabolic drugs for variable selection
meta_drugs <- c("METFORMIN","GLICLAZIDE","ATORVASTATIN", "SIMVASTATIN", "PIOGLITAZONE", "VALPROATE", "LITHIUM", "FLUOXETINE", "CITALOPRAM", "SERTRALINE", "FLUPENTHIXOL")

injectable_sga <- c("OLANZAPINE", "RISPERIDONE", "ARIPIPRAZOLE", "PALIPERIDONE")

ALL_drugs <- c(SGA_drugs, FGA_drugs, meta_drugs)

# Set the FGA indicator
df.dose$fga <- rep(0, nrow(df.dose))
df.dose[df.dose$drug_str1 %in% FGA_drugs, ncol(df.dose)] <- 1

```

### Extract the metabolic and clinical data of all subjects from dataset with medication prescription.
```{r, results='hide'}
library(lubridate)

# Calculate the duration of treatment (in months, baseline = 0)
calc_tx_duration <- function(i, df){
  df.tmp <- df[df$ID==i,]
  n <- nrow(df.tmp)
  if(n > 1) {
    out <- rep(0, n)
    for (j in 2:n) {
      out[j] <- (interval(ymd(df.tmp[1,c("Test_Date")]), ymd(df.tmp[j, c("Test_Date")]))
                 / days(30))
    }
    return(out)
  } else { # patient with only one measurement record
    return(c(0))
  }
}


# Define the response (y) and covariates (x)
x <- c("ID", "Gender", "DOB", "Onset_Age", "education", "Date", "BL_Date", "BGF_Date")
y <- c("BMI", "BGF", "cholestrol", "LDL_Chol", "HDL_Chol", "Triglycerides")

tt <- c("BL", "BGF", "BMI")
clinics <- list()

# Extract records with non-NA metabolic data
clinics[[ tt[1] ]] <- subset(df.clinic, !is.na(BL_Date))
clinics[[ tt[2] ]] <- subset(df.clinic, !is.na(BGF_Date))
clinics[[ tt[3] ]] <- subset(df.clinic, !is.na(Date))
clinics[[ tt[3] ]] <- subset(clinics[[ tt[3] ]], !is.na(BMI))

# Extract columns with subject's clinical and metabolic data
clinics[[ tt[1] ]] <- subset(clinics[[ tt[1] ]], select=c(x[-c(6,8)], y[-c(1:2)]))
clinics[[ tt[2] ]] <- subset(clinics[[ tt[2] ]], select=c(x[-c(6,7)], y[2]))
clinics[[ tt[3] ]] <- subset(clinics[[ tt[3] ]], select=c(x[-c(7:8)], y[1]))

for (t in tt){
  colnames(clinics[[t]])[6] <- "Test_Date"

  # Convert to Date
  clinics[[t]]$Test_Date <- as.Date(clinics[[t]]$Test_Date, "%d/%m/%Y")
  clinics[[t]]$DOB <- as.Date(clinics[[t]]$DOB, "%d/%m/%Y")

  # Calculate age at measurement
  clinics[[t]]$age <- interval(clinics[[t]]$DOB,
                                    clinics[[t]]$Test_Date)/days(365)

  # Sorted by ID and Test_Date
  clinics[[t]] <- clinics[[t]][with(clinics[[t]], order(ID, Test_Date)),]

  # Calculate the treatment duration for blood test data set
  ids <- unique(clinics[[t]]$ID)
  tx_duration <- sapply(ids, calc_tx_duration, clinics[[t]])
  clinics[[t]]$tx_duration <- unlist(tx_duration)

  # De-duplicate the df.clinic.bt and df.clinic.bc
  clinics[[t]] <- unique(clinics[[t]])
}
```


### Define a function to extract drug prescribed n (=windowSize) days before the metabolic level measurements.
```{r, results='hide'}
library(zoo)

# Functions definition

Modes <- function(x) {
  ux <- unique(x)
  tab <- tabulate(match(x, ux))
  ux[tab == max(tab)]
}

extractDose <- function(id, windowSize, df.clinic, df.dosage, drugNames) {
  df <- subset(df.dosage, ID==id)
  df.clinic.1 <- df.clinic[df.clinic$ID==id,]
  testdate <- df.clinic.1[, c("Test_Date")]
  
  df$startdate <- as.Date(df$startdate, "%m/%d/%Y")
  df$enddate <- as.Date(df$enddate, "%m/%d/%Y")

  n <- length(testdate)
  n.drugs <- length(drugNames)
  sga <- rep(0, n)
  fga <- rep(0, n)
  sga_tablet_cnt <- rep(0, n)
  fga_tablet_cnt <- rep(0, n)
  patient_type <- rep(NA, n)
  df.drugs <- data.frame(matrix(ncol=n.drugs, nrow=0))
  for(i in 1:n){

    # Find those drug dispense records with start_date earlier than test date by windowSize
    drugStartDate <- testdate[i] - days(windowSize)
    df.tmp <- df[drugStartDate %within% interval(df$startdate, df$enddate),]

    # Filter away the optional drugs
    df.tmp <- df.tmp[!is.na(df.tmp$total_dose),]

    # Skip to next iteration if no dispense records are found during the specific period
    if(nrow(df.tmp) == 0) {
      drugs.row <- rep(0, n.drugs)
      df.drugs <- rbind(df.drugs, drugs.row)
      next
    }
    
    # Construct a dataframe for drug details
    drugsTaken <- unique(df.tmp$drug_str1)
    drugs.idx <- which(drugNames %in% drugsTaken)
    drugs.row <- rep(0, n.drugs)
    drugs.row[drugs.idx] <- 1
    df.drugs <- rbind(df.drugs, drugs.row)

    # Calculate the total daily dosage for SGA and FGA drug in the unit of tablet
    df.tmp$tablet_cnt <- df.tmp$total_dose/df.tmp$drug_mg
    df.tmp <- df.tmp[,c("ID", "Patient.Type", "drug_str1", "sga", "fga", "tablet_cnt")]
    
    df.min_cnt <- aggregate(tablet_cnt ~ drug_str1, data=df.tmp, min)
    df.tmp <- merge(df.tmp, df.min_cnt, by="drug_str1", all.x=TRUE)[,-6]
    df.tmp <- unique(df.tmp)
    colnames(df.tmp)[6] <- "tablet_cnt"
    sga_cnts <- aggregate(tablet_cnt ~ sga, data=df.tmp, sum)
    fga_cnts <- aggregate(tablet_cnt ~ fga, data=df.tmp, sum)

    # Extract the patient type (in/out patient)
    patient_type[i] <- df.tmp$Patient.Type[1]
    if(sum(sga_cnts$sga) > 0) {
      sga[i] <- 1
      sga_tablet_cnt[i] <- sga_cnts[sga_cnts$sga==1, c("tablet_cnt")]
      
    }
    # Extract the FGA count
    if(sum(fga_cnts$fga) > 0) {
      fga[i] <- 1
      fga_tablet_cnt[i] <- fga_cnts[fga_cnts$fga==1, c("tablet_cnt")]
    }
  } # End of for-loop
  
  # Fill the NA value of Patient Type by the Mode in the vector
  fill.val <- Modes(patient_type)
  patient_type <- na.fill(patient_type, fill.val)
  
  # Merge the extracted dosage with the input clinical data frame
  df.dose.1 <- data.frame(id, testdate, patient_type, 
                          sga, fga, sga_tablet_cnt, fga_tablet_cnt)
  colnames(df.dose.1)[1:3] <- c("ID", "Test_Date", "Patient_Type")
  df.out <- merge(df.clinic.1, df.dose.1, by=c("ID", "Test_Date"), all.x=TRUE)

  # Combine the clinic and drug info dataframe
  colnames(df.drugs) <- drugNames
  df.out <- cbind(df.out, df.drugs)

  return(df.out)
}


```


#*********************************
# Disaggregation of the between-subject and within-subject effects
# newer method: ie calculating regression intercept and residuals from 
# the mean in each subject
#**********************************

```{r}
# Estimate the intercept and residual of linear regression model for between-subject and
# within-subject component respectively
calc_covar_intercept_residuals <- function(outcome, covar, df){
  ids <- unique(df$ID)
  n <- nrow(df)
  
  intercept <- rep(NA, n)
  residual <- rep(NA, n)
  for(i in 1:length(ids)){
    ind <- which(df$ID %in% ids[i])
    df.tmp <- subset(df, select=c(outcome, covar))
    df.tmp <- df.tmp[ind,]
    
    if(nrow(df.tmp) >= 3){
      n_unique <- length(unique(df.tmp[[outcome]]))
      if(n_unique >= 2){
        model <- lm(as.formula(paste0(outcome, "~", covar)), df.tmp)
        intercept[ind] <- summary(model)$coef[1,1]
        residual[ind] <- model$residual
      }
    }
    if(sum(is.na(intercept[ind])) > 0){
      outcome_mean <- mean(df.tmp[[outcome]])
      intercept[ind] <- outcome_mean
      residual[ind] <- df.tmp[[outcome]] - outcome_mean
    }
  }
  return(list(intercept, residual))
}

# Return a dataframe with the between-subject and within-subject components of a 
# independent variables in the LMM model
add_bz_wz_cols <- function(drugs, t_covariates, df) {
  results <- lapply(drugs, calc_covar_intercept_residuals, t_covariates, df)
  df.out <- df
  for(i in 1:length(drugs)){
    bz_wz <- results[[i]]
    df.out <- cbind(df.out, bz_wz[[1]], bz_wz[[2]])
    
    n_col <- ncol(df.out)
    colnames(df.out)[n_col-1] <- paste0("bz_", drugs[i])
    colnames(df.out)[n_col] <- paste0("wz_", drugs[i])
  }
  return(df.out)
}


```


### Read and integrate the drug dosage
```{r}
VERSION="paper20231025"
BATCHID="CPH.20211217"
DATAPATH=paste0("/home/kenneth/data/metabolic/ver20220920/")
OUTPATH=paste0("/home/kenneth/out/metabolic/", BATCHID, "/", VERSION, "/phenotype/")

# Conversion between injection and oral dose.
inject.to.oral.dose <- list("OLANZAPINE"=c("0"=0, "10"=10),
                       "RISPERIDONE"=c("0"=0, "5"=4 ,"25"=2, "50"=5),
                       "ARIPIPRAZOLE"=c("0"=0, "300"=10,  "400"=10),
                       "PALIPERIDONE"=c("0"=0, "525"=9, "150"=9, "100"=6, "75"=3, "50"=3))

conv_dose <- function(dose, lookup_dose) {
  lookup_dose[[as.character(dose)]]
}

# Conversion from injection dose to oral dose
# Refer to conversion table in https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9645288/
injectToOralDosage <- function (obs_inject_sga, df_dose, lookup_dose, oral_col_prefix) {

  out <- list()
  for (sga in obs_inject_sga) {
    oral_col <- paste0(oral_col_prefix, sga)
    inject_col <- paste0(injectable_sga_col_prefix, sga)
    
    inject_dose <- sapply(df_dose[[inject_col]], conv_dose, lookup_dose[[sga]])
    ttl_dose <- df_dose[[oral_col]] + inject_dose
    
    out <- append(out, list(ttl_dose))
  }
  names(out) <- obs_inject_sga
  return(out)
}


# Define the columns to be included in final output dataset
col_prefix <- "daily_dose_before.21_days.oral_."
injectable_sga_col_prefix <- "daily_dose_before.21_days.injection_."

oral_cols <- paste0(col_prefix, ALL_drugs)
inject_cols <- paste0(injectable_sga_col_prefix, injectable_sga)

traits <- c("BL", "BGF", "BMI")

# Read the dosage dataset
df.dose.list <- list()
for(t in traits){
  df.dose.list[[t]] <- read.csv(paste0(DATAPATH, "medication_dose_calculated_", t ,".csv"))
  
  # Injection dose convert to oral dose
  inject_col_idx <- match(inject_cols, colnames(df.dose.list[[t]]))
  obs_inject_sga <- injectable_sga[!is.na(inject_col_idx)]
  
  list.ttl_dose <- injectToOralDosage(obs_inject_sga, 
                                      df.dose.list[[t]], 
                                      inject.to.oral.dose, 
                                      col_prefix)
  
  # Replace the oral dose by total dose (oral + injection)
  for (sga in names(list.ttl_dose)) {
    dose_col <- paste0(col_prefix, sga)
    df.dose.list[[t]][[dose_col]] <- list.ttl_dose[[sga]]
  }
  
  # Select the dose columns

  select_cols <- c("ID", "Date_measurement", paste0(col_prefix, ALL_drugs))
  drug_idx <- match(select_cols, colnames(df.dose.list[[t]]))
  drug_idx <- drug_idx[!is.na(drug_idx)]
  
  # Remove the drugs without prescription records in those patients
  df.dose.list[[t]] <- df.dose.list[[t]][,drug_idx]

  # Rename the columns
  new_col_prefix <- "dose."
  colnames(df.dose.list[[t]]) <- sub(col_prefix, new_col_prefix, colnames(df.dose.list[[t]]))
  colnames(df.dose.list[[t]])[2] <- "Test_Date"

  # Convert the date format
  df.dose.list[[t]]$Test_Date <- as.Date(df.dose.list[[t]]$Test_Date, "%Y-%m-%d")
}
```


### Generate the final dataset suitable for building the linear mixed effect model
```{r, results='hide', message=FALSE}
library(data.table)

windowSize = 21

# Extract all drug names
drugNames <- unique(df.dose$drug_str1)

# Define the extract field idx
indexes <- list(BL=c(1:3,11:13,5:10,14:567),
                BGF=c(1:3,8:10,5:7,11:564),
                BMI=c(1:3,8:10,5:7,11:564))

# Populate the dataset for blood test related metabolic data
for (t in traits){
  ids = unique(clinics[[t]]$ID)
  #for (i in ids[585:740]) {
  #  print(i)
  #  extractDose(i, windowSize, df.clinic.bt, df.dose, drugNames)
  #}
  list.df <- lapply(ids, extractDose, windowSize, clinics[[t]], df.dose, drugNames)
  df.final <- rbindlist(list.df)
  df.final <- subset(df.final, select=colnames(df.final)[indexes[[t]]])

  # Convert the character to numeric type
  df.final$Gender <- as.factor(df.final$Gender) # 1=Male, 2=Female
  df.final$Patient_Type <- as.factor(df.final$Patient_Type)

  # Merge with the calculated dosage
  df.final.out <- merge(df.final, df.dose.list[[t]], 
                           by=c("ID", "Test_Date"), all.x=TRUE)

  # Set the dosage to 0 for the first measurement record
  first_dose_idx <- match("dose.CLOZAPINE", colnames(df.final.out))
  df.tmp <- df.final.out[,first_dose_idx:ncol(df.final.out)]
  df.tmp[is.na(df.tmp)] <- 0
  df.final.out[, first_dose_idx:ncol(df.final.out)] <- df.tmp
  rm(df.tmp)

  # Dis-aggregate time-covariates to between-subject (bz) and 
  # within-subject (wz) covariates
  t_covariates="tx_duration"
  drugs_with_dose <- colnames(df.final.out)[first_dose_idx:ncol(df.final.out)]
  wb_drugs <- unique(c("sga", "fga", ALL_drugs, drugs_with_dose))
  df.out <- add_bz_wz_cols(wb_drugs, t_covariates, df.final.out)
  
  # Output the pre-processed dataset to file
  write.table(df.out, col.names=TRUE, row.names=FALSE, quote=FALSE, sep=",",
              file=paste0(OUTPATH, "metabolic_with_total_dose_", 
                          windowSize, "days.", t, ".csv"))
}


```