---
title: "Building linear mixed effect models"
author: "Kenneth Wong Chi Yin"
date: "06/05/2024"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Model building
```{r results='hide', warning=FALSE, message=FALSE}
VERSION="paper20231025"
BATCHID="CPH.20211217"
suffix="rm_outliers.csv"
DATAPATH=paste0("/home/kenneth/out/metabolic/", BATCHID, "/", VERSION, "/phenotype/")
OUTPATH=paste0("/home/kenneth/out/metabolic/", BATCHID, "/", VERSION, "/phenotype/")

windowSize=21

# Read the Blood Test dataset 
df.bt.final <- read.csv(paste0(DATAPATH, "metabolic_with_total_dose_", windowSize, "days.BL.", suffix))

# Read the Body Check dataset
df.bc.final <- read.csv(paste0(DATAPATH, "metabolic_with_total_dose_", windowSize, "days.BMI.", suffix))

# Read the Blood Glucose Test dataset 
df.bg.final <- read.csv(paste0(DATAPATH, "metabolic_with_total_dose_", windowSize, "days.BGF.", suffix))

# Classification of drugs
# Candidate SGAs for variable selection
SGA_drugs <- c("CLOZAPINE","OLANZAPINE","ARIPIPRAZOLE","AMISULPRIDE","PALIPERIDONE","RISPERIDONE","QUETIAPINE", "ZIPRASIDONE", "LURASIDONE")

# Candidate FGAs for variable selection
FGA_drugs <- c("CHLORPROMAZINE","FLUPHENAZINE", "HALOPERIDOL", "PERPHENAZINE", "TRIFLUOPERAZINE", "PROCHLORPERAZINE", "PIMOZIDE")

# Candidate metabolic and other drugs for variable selection
meta_lipid_drugs <- c("METFORMIN", "ATORVASTATIN", "SIMVASTATIN")
other_drugs <- c("VALPROATE", "LITHIUM", "CITALOPRAM", "SERTRALINE")

# The metabolic measurement (i.e. the outcomes of LMM model)
metabolic_params <- c("cholestrol", "LDL_Chol", "HDL_Chol", "Triglycerides", "BGF", "BMI")

# Determine the 1st column index of drug dose.
first_dose_idx <- match("bz_dose.CLOZAPINE", colnames(df.bt.final))
drugs_with_dose <- colnames(df.bt.final)[first_dose_idx:ncol(df.bt.final)]


```


## Function to present the coefficient of LMM model in a pretty way.
```{r echo = FALSE, warning=FALSE, message=FALSE}
library(sjPlot)

plot_lmm_model <- function(data, axis_label, title){
  sjPlot::plot_model(data[[1]],
                   axis.labels=axis_label,
                   show.values=TRUE, show.p=TRUE,
                   title=paste0(title, data[[2]]))
}

visualize_model <- function(models, model_names, label, title, tfm){

  #data <- list()
  #for(i in 1:length(model_names)){
  #  data <- c(data, list(list(models[[i]], model_names[i])))
  #}
  #plot.out <- lapply(data, plot_lmm_model, label, title)
  
  pl <- c(
    `(Intercept)` = "Intercept",
    age = "Age",
    Gender = "Sex (Female)",
    education = "Year of Education",
    Patient_TypeOP = "Out-Patient",
    sga = "Take 2nd Gen. Drug (SGA)",
    fga = "Take 1st Gen. Drug (FGA)",
    sga_tablet_cnt = "SGA Daily tablet cnt",
    fga_tablet_cnt = "FGA Daily table cnt",
    tx_duration = "Tx Duration (months) ___________________",
    btw_effect = "Btw-subj Drug effect",
    within_effect = "Within-subj Drug effect",
    bz_CLOZAPINE = "bz.CLOZAPINE",
    bz_OLANZAPINE = "bz.OLANZAPINE",
    bz_ARIPIPRAZOLE = "bz.ARIPIPRAZOLE",
    bz_AMISULPRIDE = "bz.AMISULPRIDE",
    bz_QUETIAPINE = "bz.QUETIAPINE",
    bz_RISPERIDONE = "bz.RISPERIDONE",
    bz_PALIPERIDONE = "bz.PALIPERIDONE",
    bz_METFORMIN = "bz.METFORMIN",
    bz_ATORVASTATIN = "bz.ATORVASTATIN",
    bz_SIMVASTATIN = "bz.SIMVASTATIN",
    bz_VALPROATE = "bz.VALPROATE",
    bz_LITHIUM = "bz.LITHIUM",
    bz_CITALOPRAM = "bz.CITALOPRAM",
    bz_SERTRALINE = "bz.SERTRALINE ___________________",
    bz_dose.SERTRALINE = "bz_dose.SERTRALINE ___________________",
    wz_CLOZAPINE = "wz.CLOZAPINE",
    wz_OLANZAPINE = "wz.OLANZAPINE",
    wz_ARIPIPRAZOLE = "wz.ARIPIPRAZOLE",
    wz_AMISULPRIDE = "wz.AMISULPRIDE",
    wz_QUETIAPINE = "wz.QUETIAPINE",
    wz_RISPERIDONE = "wz.RISPERIDONE",
    wz_PALIPERIDONE = "wz.PALIPERIDONE",
    wz_METFORMIN = "wz.METFORMIN",
    wz_ATORVASTATIN = "wz.ATORVASTATIN",
    wz_SIMVASTATIN = "wz.SIMVASTATIN",
    wz_VALPROATE = "wz.VALPROATE",
    wz_LITHIUM = "wz.LITHIUM",
    wz_CITALOPRAM = "wz.CITALOPRAM",
    wz_SERTRALINE = "wz.SERTRALINE",
    a_non_used_label = "We don't care"
  )
  
  #summary <- sjPlot::tab_model(models, pred.labels=pl, dv.labels=model_names, title=title,
  #                             show.ci=FALSE, show.p=TRUE, show.r2=TRUE, transform=tfm,
  #                             digits=4, digits.re=4)
  summary <- sjPlot::tab_model(models, pred.labels=pl, dv.labels=model_names, title=title,
                               show.stat=TRUE, show.se=TRUE, show.p=TRUE, show.r2=TRUE, transform=tfm,
                               digits=4, digits.re=4)  
  print(summary)
  #return(plot.out)
}
```


## Building Linear Mixed Model to estimate the Between- and Within- Subject effect of all
## SGA drugs (ignore dosage, use binary 0/1 to indicate drug taken or not)
```{r include=TRUE, warning=TRUE, message=TRUE}
library(lme4)

### Model for Cholestrol ###
df.bt.final.tmp <- df.bt.final

# Transform the metobolic level
df.bt.final.tmp$log_cholestrol <- log(df.bt.final$cholestrol)

TC_mixed_model_binary <- lmer(log_cholestrol~age+Gender+education+tx_duration+bz_CLOZAPINE+bz_OLANZAPINE+bz_ARIPIPRAZOLE+bz_AMISULPRIDE+bz_QUETIAPINE+bz_RISPERIDONE+bz_PALIPERIDONE+bz_METFORMIN+bz_ATORVASTATIN+bz_SIMVASTATIN+bz_VALPROATE+bz_LITHIUM+bz_CITALOPRAM+bz_SERTRALINE
                         +wz_CLOZAPINE+wz_OLANZAPINE+wz_ARIPIPRAZOLE+wz_AMISULPRIDE+wz_QUETIAPINE+wz_RISPERIDONE+wz_PALIPERIDONE+wz_METFORMIN+wz_ATORVASTATIN+wz_SIMVASTATIN+wz_VALPROATE+wz_LITHIUM+wz_CITALOPRAM+wz_SERTRALINE
                         +(1|ID), df.bt.final.tmp)
summary(TC_mixed_model_binary)

### Model for HDL_Chol ###
df.bt.final.tmp$log_HDL_Chol <- log(df.bt.final$HDL_Chol)

HDL_mixed_model_binary <- lmer(log_HDL_Chol~age+Gender+education+tx_duration+bz_CLOZAPINE+bz_OLANZAPINE+bz_ARIPIPRAZOLE+bz_AMISULPRIDE+bz_QUETIAPINE+bz_RISPERIDONE+bz_PALIPERIDONE+bz_METFORMIN+bz_ATORVASTATIN+bz_SIMVASTATIN+bz_VALPROATE+bz_LITHIUM+bz_CITALOPRAM+bz_SERTRALINE
                         +wz_CLOZAPINE+wz_OLANZAPINE+wz_ARIPIPRAZOLE+wz_AMISULPRIDE+wz_QUETIAPINE+wz_RISPERIDONE+wz_PALIPERIDONE+wz_METFORMIN+wz_ATORVASTATIN+wz_SIMVASTATIN+wz_VALPROATE+wz_LITHIUM+wz_CITALOPRAM+wz_SERTRALINE
                         +(1|ID), df.bt.final.tmp)
summary(HDL_mixed_model_binary)

### Model for LDL ###
df.bt.final.tmp$log_LDL_Chol <- log(df.bt.final$LDL_Chol)

LDL_mixed_model_binary <- lmer(log_LDL_Chol~age+Gender+education+tx_duration+bz_CLOZAPINE+bz_OLANZAPINE+bz_ARIPIPRAZOLE+bz_AMISULPRIDE+bz_QUETIAPINE+bz_RISPERIDONE+bz_PALIPERIDONE+bz_METFORMIN+bz_ATORVASTATIN+bz_SIMVASTATIN+bz_VALPROATE+bz_LITHIUM+bz_CITALOPRAM+bz_SERTRALINE
                         +wz_CLOZAPINE+wz_OLANZAPINE+wz_ARIPIPRAZOLE+wz_AMISULPRIDE+wz_QUETIAPINE+wz_RISPERIDONE+wz_PALIPERIDONE+wz_METFORMIN+wz_ATORVASTATIN+wz_SIMVASTATIN+wz_VALPROATE+wz_LITHIUM+wz_CITALOPRAM+wz_SERTRALINE
                         +(1|ID), df.bt.final.tmp)
summary(LDL_mixed_model_binary)


### Model for Triglycerides ###
#df.bt.final.tmp$sqrt_Triglycerides <- (1/df.bt.final$Triglycerides)^(1/4)
df.bt.final.tmp$log_Triglycerides <- log(df.bt.final$Triglycerides)

TG_mixed_model_binary <- lmer(log_Triglycerides~age+Gender+education+tx_duration+bz_CLOZAPINE+bz_OLANZAPINE+bz_ARIPIPRAZOLE+bz_AMISULPRIDE+bz_QUETIAPINE+bz_RISPERIDONE+bz_PALIPERIDONE+bz_METFORMIN+bz_ATORVASTATIN+bz_SIMVASTATIN+bz_VALPROATE+bz_LITHIUM+bz_CITALOPRAM+bz_SERTRALINE
                         +wz_CLOZAPINE+wz_OLANZAPINE+wz_ARIPIPRAZOLE+wz_AMISULPRIDE+wz_QUETIAPINE+wz_RISPERIDONE+wz_PALIPERIDONE+wz_METFORMIN+wz_ATORVASTATIN+wz_SIMVASTATIN+wz_VALPROATE+wz_LITHIUM+wz_CITALOPRAM+wz_SERTRALINE
                         +(1|ID), df.bt.final.tmp)
summary(TG_mixed_model_binary)


### Model for BGF ###
df.bg.final.tmp <- df.bg.final
df.bg.final.tmp$log_BGF <- log(df.bg.final$BGF)

BGF_mixed_model_binary <- lmer(log_BGF~age+Gender+education+tx_duration+bz_CLOZAPINE+bz_OLANZAPINE+bz_ARIPIPRAZOLE+bz_AMISULPRIDE+bz_QUETIAPINE+bz_RISPERIDONE+bz_PALIPERIDONE+bz_METFORMIN+bz_ATORVASTATIN+bz_SIMVASTATIN+bz_VALPROATE+bz_LITHIUM+bz_CITALOPRAM+bz_SERTRALINE
                         +wz_CLOZAPINE+wz_OLANZAPINE+wz_ARIPIPRAZOLE+wz_AMISULPRIDE+wz_QUETIAPINE+wz_RISPERIDONE+wz_PALIPERIDONE+wz_METFORMIN+wz_ATORVASTATIN+wz_SIMVASTATIN+wz_VALPROATE+wz_LITHIUM+wz_CITALOPRAM+wz_SERTRALINE
                         +(1|ID), df.bg.final.tmp)
summary(BGF_mixed_model_binary)


### Model for BMI ###
df.bc.final.tmp <- df.bc.final
df.bc.final.tmp$log_BMI <- log(df.bc.final$BMI)

BMI_mixed_model_binary <- lmer(log_BMI~age+Gender+education+tx_duration+bz_CLOZAPINE+bz_OLANZAPINE+bz_ARIPIPRAZOLE+bz_AMISULPRIDE+bz_QUETIAPINE+bz_RISPERIDONE+bz_PALIPERIDONE+bz_METFORMIN+bz_ATORVASTATIN+bz_SIMVASTATIN+bz_VALPROATE+bz_LITHIUM+bz_CITALOPRAM+bz_SERTRALINE
                         +wz_CLOZAPINE+wz_OLANZAPINE+wz_ARIPIPRAZOLE+wz_AMISULPRIDE+wz_QUETIAPINE+wz_RISPERIDONE+wz_PALIPERIDONE+wz_METFORMIN+wz_ATORVASTATIN+wz_SIMVASTATIN+wz_VALPROATE+wz_LITHIUM+wz_CITALOPRAM+wz_SERTRALINE
                         +(1|ID), df.bc.final.tmp)
summary(BMI_mixed_model_binary)

#save(TC_mixed_model_binary, HDL_mixed_model_binary, LDL_mixed_model_binary, TG_mixed_model_binary, BGF_mixed_model_binary, BMI_mixed_model_binary, 
#     file=file.path(paste0("/home/kenneth/out/metabolic/", BATCHID, "/", VERSION, "/MI/lmm.binary.raw.Rdata")))
```

## Building Linear Mixed Model to estimate the Between- and Within- Subject effect of all SGA drugs with dosage

```{r include=TRUE, warning=TRUE, message=TRUE}
library(lme4)

### Model for Cholestrol ###
df.bt.final.tmp <- df.bt.final

# Transform the metobolic level
df.bt.final.tmp$log_cholestrol <- log(df.bt.final$cholestrol)

TC_mixed_model_dose <- lmer(log_cholestrol~age+Gender+education+tx_duration+bz_dose.CLOZAPINE+bz_dose.OLANZAPINE+bz_dose.ARIPIPRAZOLE+bz_dose.AMISULPRIDE+bz_dose.QUETIAPINE+bz_dose.RISPERIDONE+bz_dose.PALIPERIDONE+bz_dose.METFORMIN+bz_dose.ATORVASTATIN+bz_dose.SIMVASTATIN+bz_dose.VALPROATE+bz_dose.LITHIUM+bz_dose.CITALOPRAM+bz_dose.SERTRALINE
                         +wz_dose.CLOZAPINE+wz_dose.OLANZAPINE+wz_dose.ARIPIPRAZOLE+wz_dose.AMISULPRIDE+wz_dose.QUETIAPINE+wz_dose.RISPERIDONE+wz_dose.PALIPERIDONE+wz_dose.METFORMIN+wz_dose.ATORVASTATIN+wz_dose.SIMVASTATIN+wz_dose.VALPROATE+wz_dose.LITHIUM+wz_dose.CITALOPRAM+wz_dose.SERTRALINE
                         +(1|ID), df.bt.final.tmp)
summary(TC_mixed_model_dose)


### Model for HDL_Chol ###
df.bt.final.tmp$log_HDL_Chol <- log(df.bt.final$HDL_Chol)

HDL_mixed_model_dose <- lmer(log_HDL_Chol~age+Gender+education+tx_duration+bz_dose.CLOZAPINE+bz_dose.OLANZAPINE+bz_dose.ARIPIPRAZOLE+bz_dose.AMISULPRIDE+bz_dose.QUETIAPINE+bz_dose.RISPERIDONE+bz_dose.PALIPERIDONE+bz_dose.METFORMIN+bz_dose.ATORVASTATIN+bz_dose.SIMVASTATIN+bz_dose.VALPROATE+bz_dose.LITHIUM+bz_dose.CITALOPRAM+bz_dose.SERTRALINE
                         +wz_dose.CLOZAPINE+wz_dose.OLANZAPINE+wz_dose.ARIPIPRAZOLE+wz_dose.AMISULPRIDE+wz_dose.QUETIAPINE+wz_dose.RISPERIDONE+wz_dose.PALIPERIDONE+wz_dose.METFORMIN+wz_dose.ATORVASTATIN+wz_dose.SIMVASTATIN+wz_dose.VALPROATE+wz_dose.LITHIUM+wz_dose.CITALOPRAM+wz_dose.SERTRALINE
                         +(1|ID), df.bt.final.tmp)
summary(HDL_mixed_model_dose)


### Model for LDL ###
df.bt.final.tmp$log_LDL_Chol <- log(df.bt.final$LDL_Chol)

LDL_mixed_model_dose <- lmer(log_LDL_Chol~age+Gender+education+tx_duration+bz_dose.CLOZAPINE+bz_dose.OLANZAPINE+bz_dose.ARIPIPRAZOLE+bz_dose.AMISULPRIDE+bz_dose.QUETIAPINE+bz_dose.RISPERIDONE+bz_dose.PALIPERIDONE+bz_dose.METFORMIN+bz_dose.ATORVASTATIN+bz_dose.SIMVASTATIN+bz_dose.VALPROATE+bz_dose.LITHIUM+bz_dose.CITALOPRAM+bz_dose.SERTRALINE
                        +wz_dose.CLOZAPINE+wz_dose.OLANZAPINE+wz_dose.ARIPIPRAZOLE+wz_dose.AMISULPRIDE+wz_dose.QUETIAPINE+wz_dose.RISPERIDONE+wz_dose.PALIPERIDONE+wz_dose.METFORMIN+wz_dose.ATORVASTATIN+wz_dose.SIMVASTATIN+wz_dose.VALPROATE+wz_dose.LITHIUM+wz_dose.CITALOPRAM+wz_dose.SERTRALINE
                         +(1|ID), df.bt.final.tmp)
summary(LDL_mixed_model_dose)


### Model for Triglycerides ###
df.bt.final.tmp$log_Triglycerides <- log(df.bt.final$Triglycerides)

TG_mixed_model_dose <- lmer(log_Triglycerides~age+Gender+education+tx_duration+bz_dose.CLOZAPINE+bz_dose.OLANZAPINE+bz_dose.ARIPIPRAZOLE+bz_dose.AMISULPRIDE+bz_dose.QUETIAPINE+bz_dose.RISPERIDONE+bz_dose.PALIPERIDONE+bz_dose.METFORMIN+bz_dose.ATORVASTATIN+bz_dose.SIMVASTATIN+bz_dose.VALPROATE+bz_dose.LITHIUM+bz_dose.CITALOPRAM+bz_dose.SERTRALINE
                        +wz_dose.CLOZAPINE+wz_dose.OLANZAPINE+wz_dose.ARIPIPRAZOLE+wz_dose.AMISULPRIDE+wz_dose.QUETIAPINE+wz_dose.RISPERIDONE+wz_dose.PALIPERIDONE+wz_dose.METFORMIN+wz_dose.ATORVASTATIN+wz_dose.SIMVASTATIN+wz_dose.VALPROATE+wz_dose.LITHIUM+wz_dose.CITALOPRAM+wz_dose.SERTRALINE
                         +(1|ID), df.bt.final.tmp)
summary(TG_mixed_model_dose)


### Model for BGF ###
df.bg.final.tmp <- df.bg.final
df.bg.final.tmp$log_BGF <- log(df.bg.final$BGF)

BGF_mixed_model_dose <- lmer(log_BGF~age+Gender+education+tx_duration+bz_dose.CLOZAPINE+bz_dose.OLANZAPINE+bz_dose.ARIPIPRAZOLE+bz_dose.AMISULPRIDE+bz_dose.QUETIAPINE+bz_dose.RISPERIDONE+bz_dose.PALIPERIDONE+bz_dose.METFORMIN+bz_dose.ATORVASTATIN+bz_dose.SIMVASTATIN+bz_dose.VALPROATE+bz_dose.LITHIUM+bz_dose.CITALOPRAM+bz_dose.SERTRALINE
                         +wz_dose.CLOZAPINE+wz_dose.OLANZAPINE+wz_dose.ARIPIPRAZOLE+wz_dose.AMISULPRIDE+wz_dose.QUETIAPINE+wz_dose.RISPERIDONE+wz_dose.PALIPERIDONE+wz_dose.METFORMIN+wz_dose.ATORVASTATIN+wz_dose.SIMVASTATIN+wz_dose.VALPROATE+wz_dose.LITHIUM+wz_dose.CITALOPRAM+wz_dose.SERTRALINE
                         +(1|ID), df.bg.final.tmp)
summary(BGF_mixed_model_dose)


### Model for BMI ###
df.bc.final.tmp <- df.bc.final
df.bc.final.tmp$log_BMI <- log(df.bc.final$BMI)

BMI_mixed_model_dose <- lmer(log_BMI~age+Gender+education+tx_duration+bz_dose.CLOZAPINE+bz_dose.OLANZAPINE+bz_dose.ARIPIPRAZOLE+bz_dose.AMISULPRIDE+bz_dose.QUETIAPINE+bz_dose.RISPERIDONE+bz_dose.PALIPERIDONE+bz_dose.METFORMIN+bz_dose.ATORVASTATIN+bz_dose.SIMVASTATIN+bz_dose.VALPROATE+bz_dose.LITHIUM+bz_dose.CITALOPRAM+bz_dose.SERTRALINE
                         +wz_dose.CLOZAPINE+wz_dose.OLANZAPINE+wz_dose.ARIPIPRAZOLE+wz_dose.AMISULPRIDE+wz_dose.QUETIAPINE+wz_dose.RISPERIDONE+wz_dose.PALIPERIDONE+wz_dose.METFORMIN+wz_dose.ATORVASTATIN+wz_dose.SIMVASTATIN+wz_dose.VALPROATE+wz_dose.LITHIUM+wz_dose.CITALOPRAM+wz_dose.SERTRALINE
                         +(1|ID), df.bc.final.tmp)
summary(BMI_mixed_model_dose)

# Output the models to file
#save(TC_mixed_model_dose, HDL_mixed_model_dose, LDL_mixed_model_dose, TG_mixed_model_dose, BGF_mixed_model_dose, BMI_mixed_model_dose, 
#     file=file.path(paste0("/home/kenneth/out/metabolic/", BATCHID, "/", VERSION, "/MI/lmm.dose.raw.Rdata")))
```


## Print the result of LMM in a nice way
```{r eval=FALSE, result='hide', warning=FALSE, message=FALSE, echo=FALSE}
BATCHID="CPH.20211217"
VERSION="paper20231025"

# Output the Effect of CLOZAPINE on metabolic parameters
m_type = c("binary_all_drugs", "dose_all_drugs", "SGA", "dose_OZP", "dose_CZP")
m_names <- c("TC", "HDL", "LDL", "TG", "BG", "BMI")


models <- list()
# All drugs model with binary dosage (0/1)
models[[ m_type[1] ]] <- list(TC_mixed_model_binary, HDL_mixed_model_binary, LDL_mixed_model_binary, TG_mixed_model_binary, BGF_mixed_model_binary, BMI_mixed_model_binary)
# All drugs model with dosage
models[[ m_type[2] ]] <- list(TC_mixed_model_dose, HDL_mixed_model_dose, LDL_mixed_model_dose, TG_mixed_model_dose, BGF_mixed_model_dose, BMI_mixed_model_dose)

# Define the reverse transform function for the model coefficients
exp2percentInc <- function(x) { (exp(x) - 1)*100 }

title <- list("Association of SGA on TC, HDL, LDL, TG, BG and BMI",
              "Effect size of SGA dose on TC, HDL, LDL, TG, BG and BMI")

# Generate the summary statistic of the model coefficients
for (i in 1:length(models)) {
  m <- models[[m_type[i]]]
  visualize_model(m, m_names, axis_label, title[i], "exp2percentInc")
}


```




